Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model

### 2014-04-15 paper_enahpc14/Introduction.tex
updates Section 2: "The COSMO-ART model system"

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of   solar   radiation,    thereby   exerting   a   direct   radiative
forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}, or  by  the
modification of  cloud properties, cloud fraction  and surface albedo,
causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and    measurement-based    assessments   \citep{Myhre-2009},    large
uncertainties    remain    in    current    estimates    of    aerosol
forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
Rosenfeld-2013, Sherwood-2013, Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  uncertainties  in  aerosol  radiative forcing  estimates,  the
research  community  is making  a  concerted  international effort  to
represent  the  underlying physical,  chemical  and aerosol  dynamical
processes through numerical chemical  transport models (CTMs), such as
ART (Aerosols  and Reactive Trace  gases), developed at  the Karlsruhe
Institute   of  Technology   (KIT)   \citep{Vogel-2009,  Bangert-2011,
Knote-2013}. This  regional scale model system is  online coupled with
the  operational weather  forecast  model COSMO  \citep{Baldauf-2011},
jointly developed by a consortium of European weather centers and used
in the climate version (COSMO-CLM)  by a wide research community.

@@

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
to  the  investigated   performance  and  energy  evaluation  methods.
Sec.~\ref{sec:3} describes different power measurement equipments used
for  conducting  this  benchmarking  study and  presents  energy-aware
techniques  employed   to  determine  the  energy   footprint  of  the
considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
from  the Exa2Green  project and  highlights areas  where improvements
will be  necessary for the subsequent  baseline refactoring.  Finally,
we  conclude with  practical guidelines  to identifying  energy saving
opportunities  and give  directions  for further  developments of  the
modeling system in Sec.~\ref{concl}.

>>>>

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of   solar   radiation,    thereby   exerting   a   direct   radiative
forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
modification of  cloud properties, cloud fraction  and surface albedo,
causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and    measurement-based    assessments   \citep{Myhre-2009},    large
uncertainties  remain  in   current  estimates  of  aerosol  radiative
forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
Rosenfeld-2013, Sherwood-2013, Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale  model system  is online  coupled with  the  operational weather
forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
consortium of European weather centers and used in the climate version
(COSMO-CLM) by a wide research community.

@@

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART and specify its  technical setup related to the investigated
performance and energy evaluation methods.  Sec.~\ref{sec:3} describes
different  power  measurement  equipments  used  for  conducting  this
benchmarking  study and presents  energy-aware techniques  employed to
determine  the   energy  footprint  of   the  considered  application.
Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
and  highlights areas  where improvements  will be  necessary  for the
subsequent baseline refactoring.   Finally, we conclude with practical
guidelines  to  identifying   energy  saving  opportunities  and  give
directions  for  further  developments   of  the  modeling  system  in
Sec.~\ref{concl}.

### 2014-04-15 paper_enahpc14/Sec2:COSMO-ART.tex
updates Section 2: "The COSMO-ART model system"

\subsection{Model Description}
\label{subsec:1.1}
The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
regional  to  continental scale  model  coupled  online  to the  COSMO
numerical weather  prediction and climate  model \citep{Baldauf-2011}.
It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
aerosol dynamics  and allows the online calculation  of reactive trace
substances    and    their    interaction    with    the    atmosphere
\citep{Knote-2011}.  Detailed  model   description  can  be  found  in
\citep{Stanelle-2010, Bangert-2012, Knote-2011, Knote-2013}.

\subsection{Model Setup}
\label{subsec:1.2}
Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected  variance for the  duration of  the Exa2Green  project. After
consultation  with  the  climate  community  to  properly  define  the
baseline,  it was  necessary to  find a  run-configuration  capable of
being recreated in all subsequent versions of the code.

Thus, we give here a brief  overview of the main features of the model
setup for  our benchmark's daily  runs.  Three-dimensional simulations
are performed over  Europe for April 13th 2010, which  is close to the
equinox and thus  brings benefits of having approximately  half day of
night  and half  day  of  sun exposure,  therefore  ensuring a  proper
activation of the chemistry cycle. We consider daily 24-hour forecasts
without  any  previous spin-up  simulations.   The calculation  domain
corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
$222\times   216$    points   with   a    horizontal   resolution   of
$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
layers.  COSMO-ART requires the following input data:

\begin{itemize}
\item  Gas phase:  Anthropogenic emissions  for different  species and
  land use data for biogenic emissions and deposition,
\item Aerosol particles: Anthropogenic emissions,
\item Mineral dust: Soil specific land use data.
\end{itemize}

The meteorological initial and  bounday conditions are obtained by the
the ECMWF  global spectral model IFS  with an update  frequency of 3h.
Boundary data  for gas-phase species are taken  from IFS-MOZART output
at   6h  temporal   resolution.   NAMELIST-input   specifying  runtime
parameters    splitted   into    several   groups    are    given   in
Table~\ref{tab:1}.   The model setup  incorporates 34  2-d and  45 3-d
fields  to  be  written  out  every  hour  and  doesn't  include  data
assimilation methods.

\begin{table}[htbf]
  \begin{center}
    \caption{}
    \label{tab:1}
    \begin{tabular}{ll}
      \hline\noalign{\smallskip} 
      \textbf{Group} & \textbf{Description} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      LMGRID & Grid domain and size parameters \\
      RUNCTL & Model run parameters \\
      TUNING & Physics and dynamics parameters \\
      DYNCTL & Adiabatic model parameters \\
      PHYCTL & Diabatic model parameters \\
      COSMO\_ART & Gases and aerosols model parameters \\
      DIACTL & Diagnostic calculations parameters \\
      SATCTL & Synthetic satellite images parameters \\
      IOCTL & I/O environment parameters \\
      GRIBIN & GRIB input parameters \\
      GRIBOUT & GRIB output parameters \\
     \noalign{\smallskip}\hline
    \end{tabular}
  \end{center}
\end{table}

This COSMO-ART version is configured with a semi Lagrangian horizontal
advection  scheme with  tricubic interpolation  and  selective filling
diffusion  option  in  combination   with  the  dynamical  core  using
Runge-Kutta time stepping \citep{}.  It  also makes use of the Kinetic
PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
ordinary  differential equations \citep{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud  feedbacks  but  doesn't include  in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterisations,   precipitation  formation   is   performed  by   a
two-moment  cloud  microphysics scheme  instead  of  a classical  bulk
microphysics scheme.

>>>>

\subsection{Model Description}
\label{subsec:1.1}
The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
regional  to  continental scale  model  coupled  online  to the  COSMO
numerical weather  prediction and climate  model \citep{Baldauf-2011}.
It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
aerosol dynamics  and allows the online calculation  of reactive trace
substances and their interaction  with the atmosphere.  Detailed model
description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
  Knote-2013}.

\subsection{Model Setup}
\label{subsec:1.2}
Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected variance for the duration of the Exa2Green project. To define
a baseline,  it was necessary  to find a run-configuration  capable of
being  recreated in  lal subsequent  versions.  Here  we give  a brief
overview of the model setup  for our performance and energy efficiency
evaluation.

Three-dimensional simulations are performed over large parts of Europe
and the Mediterranean Sea for April 13th 2010 in order to benefit from
around  half  day  of  sun  exposure and  therefore  ensure  a  proper
activation of the chemistry cycle.  The calculation domain corresponds
to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
points with  a horizontal resolution of $0.22\,^{\circ}$,  i.e.  50 km
in  both directions and  40 vertical  layers.  These  24-hour forecast
simulations  are not  preceded  by  a spin-up  phase  to build-up  the
simulated gaseous and  aerosol concentrations.  This condition entails
that at the  model initialization there is a  period of adjustment for
the model  to reach its equilibrium  state and minimize  the effect of
initial conditions for gases and aerosols on the model predictions.

The meteorological initial and  bounday conditions are obtained by the
the ECMWF  global spectral model IFS  with an update  frequency of 3h.
Boundary data  for gas-phase species are taken  from IFS-MOZART output
at 6h temporal resolution. The  model setup incorporates 34 2-d and 45
3-d  fields  to be  written  out  every hour  and  is  devoid of  data
assimilation methods.

The  considered  COSMO-ART model  system  is  configured  with a  semi
Lagrangian horizontal advection scheme with tricubic interpolation and
selective filling  diffusion option in combination  with the dynamical
core  using Runge-Kutta  time  stepping \citep{COSMO-PartI-2011}.   It
also  makes use  of  the  Kinetic PreProcessor  solver  (KPP) for  the
resolution  of atmospheric  chemistry ordinary  differential equations
\citep{Damian-2002}.   Concerning the modelling  of wet  deposition in
aerosols, the  baseline has only indirect cloud  feedbacks but doesn't
include  in-cloud  scavenging  (rainout)  and  below-cloud  scavenging
(washout)  yet.   Amongst  physical  parameterisations,  precipitation
formation is performed by a two-moment cloud microphysics.
